Symbolic Abstract Data Type Inference

### 2015-07-08 algorithm.tex
Added back the section label that Constantin deleted. * He is so clumsy with version control!
Section titles + restructuring the restructuring.
Tried to clean up the f**king mess of the algorithm section.

%!TEX root = draft.tex
\section{Solving the Symbolic ADT Inference Problem}

\begin{figure}
\begin{lstlisting}
Min := $\emptyset$
while (true) {
  $B_i$ := $ \set{ h \in B : \weight(h) = i }$;
  MinTemp := Min $\cup$ Min($B_i$)
  if ( MinTemp = Min )
    break;
  Min := MinTemp
}
return $F$(Min);
\end{lstlisting}
\caption{The algorithm solving the symbolic ADT inference problem for a given ADT $A$, where $B$ denotes the bounded complement of $\ker A$. The procedure $\texttt{Min}(X)$ returns the $\preceq$-minimal elements of the set of histories $X$ and $\texttt{Formula}(X)$ returns a history formula which holds exactly for the histories in the upward closure of $X$.} 
\label{fig:algorithm}
\end{figure}

Figure~\ref{fig:algorithm} lists the algorithm for solving the symbolic ADT inference problem which is based on the relation $\preceq$ 
defined in Section~\ref{sec:patterns}.
%For a given set of elements $X$ and an order relation $\preceq$, the \emph{upward closure}
%of $X$ w.r.t. $\preceq$ is the set $\uparrow X=\set{y : \exists x\in X.\ x\preceq y}$. 
The algorithm performs an weight-increasing enumeration of $B$ while maintaining the
$\preceq$-minimal elements, and stops whenever each history of weight $i$ is bigger (w.r.t. $\preceq$) than some
element of the collected set of minimals.

The partial correctness of the algorithm (i.e., ignoring termination) relies on the input ADT being \emph{predictable}, i.e.,
for all $i$,
\begin{align*}
B_{i+1} \succeq \texttt{Min}(B_{\leq i}) \mbox{ implies } \texttt{Min}(B) = \texttt{Min}(B_{\leq i}),
\end{align*}
where $B_{\leq i}=\bigcup_{1\leq j\leq i}B_j$.
Section~\ref{sec:nature} shows that many naturally-occurring ADTs are predictable.
%Although we don't provide a general result concerning predictability, we prove that it holds for  (see ).

\begin{theorem}\label{th:corr1}
For any predictable ADT $A$, if the algorithm in Figure~\ref{fig:algorithm} terminates, then the returned formula represents $A$.
\end{theorem}
\begin{proof}
%For a set $\Sigma$, let $\neg \Sigma$ denote its complement.
Let $M$ be the value of \texttt{Min} at the end of the \texttt{while} loop.
%By Lemma~\ref{lem:ADT_closure}, the complement of $A$ is closed under $\preceq$. Since 
%$M\subseteq B\subseteq  \neg \ker A \subseteq \neg A$, the closure of $M$, denoted by $Cl(M)$, satisfies $Cl(M)\subseteq \neg A$. 
Since $A$ is predictable, we have that $B \subseteq Cl(M)$. % where $Cl(M)$ denotes the closure of 
Therefore, by Lemma~\ref{lem:formula1}, the returned formula represents $A$.
\end{proof}

\begin{remark}
Note that for non-predictable ADTs, the value of $\texttt{Min}$ at the end of the \texttt{while} loop 
is an under-approximation of $\texttt{Min}(B)$. 
The resulting symbolic ADT representation is still satisfied by all histories in $\ker A$, therefore it would still be sound in only identifying
\emph{actual} violations, yet incomplete in identifying \emph{all} violations.
\end{remark}

The termination of the algorithm in Figure~\ref{fig:algorithm} is a direct consequence of Lemma~\ref{lem:wqo}.
Also, since the number of weight-$i$ histories is finite\footnote{As
noted in Section~\ref{sec:histories}, we consider equality between histories up
to renaming of operation identifiers.} for each $i \in \mathbb{N}$, each $B_i$
is computable.

\begin{theorem}\label{th:corr2}
The algorithm in Figure~\ref{fig:algorithm} terminates.
\end{theorem}

>>>>

%!TEX root = draft.tex
\section{A Symbolic ADT Inference Algorithm}
\label{sec:algorithm}

Algorithm~\ref{alg:inference} lists the algorithm for solving the symbolic ADT
inference problem which is based on the relation $\preceq$ defined in
Section~\ref{sec:patterns}. The algorithm performs an weight-increasing
enumeration of $B$ while maintaining the $\preceq$-minimal elements, and stops
whenever no new minimals are found at a given weight.

\begin{algorithm}[t]
  \SetAlgoLined
  \SetKwInOut{Input}{Input}
  \Input{A reference implementation $\mathcal{I}$}
  \KwResult{A formula representing the ADT of $\mathcal{I}$}
  weight $\gets$ $0$ \;
  patterns $\gets$ $\emptyset$ \;
  \Repeat{$B_\text{weight} = B_{\text{weight}-1}$}{
    weight $\gets$ weight + $1$ \;
    patterns $\gets$ $\min B_\text{weight}$ \;
  }
  \Return $F(\text{patterns})$
  \caption{Symbolic ADT inference.}
  \label{alg:inference}
\end{algorithm}

The partial correctness of Algorithm~\ref{alg:inference} (i.e.,~ignoring
termination) relies on the input ADT being \emph{predictable}, i.e.,~for all
$i \in \mathbb{N}$,
\begin{align*}
  B_{i+1} \succeq \min \bigcup_{0 < j \le i} b_j
  \text{ implies } \min B = \min \bigcup_{0 < j \le i} b_j
\end{align*}
Section~\ref{sec:nature} shows that many naturally-occurring ADTs are predictable.

\begin{theorem}
  \label{thm:algorithm}

  Algorithm~\ref{alg:inference} terminates. If the input implementation’s
  ADT is predictable, then the returned formula represents it.

\end{theorem}

\begin{proof}
  Termination is a direct consequence of Lemma~\ref{lem:wqo}, and since the
  number of histories of any weight $i \in \mathbb{N}$ is finite\footnote{
    As noted in Section~\ref{sec:histories}, we consider equality between
    histories up to renaming of operation identifiers.
  }
  each $B_i$ is computable.

  Furthermore, when $A$ is predictable, we
  have $B \subseteq \text{patterns}^+$, where $\text{patterns}^+$ is the
  upward closure of $\text{patterns}$ under $\preceq$. Thus by
  Lemma~\ref{lem:formula1}, the returned formula represents $A$.
\end{proof}

Note that for non-predictable ADTs, the value of $\text{patterns}^+$ upon
termination of Algorithm~\ref{alg:inference} is an under-approximation of the
bounded ADT complement $B$. The resulting symbolic ADT representation is still
satisfied by all histories in $A$, therefore it would still be sound for modular
program reasoning, identifying only actual violations, and implying the
correctness of client programs which do not depend on the stronger criteria
which excludes unidentified violations. However, the under-approximation may be
incomplete in identifying all violations, and in proving client programs which
depend on the stronger criteria which excludes them all.

### 2015-07-08 inference.tex
Section titles + restructuring the restructuring.
Better overview.
Tried to clean up the f**king mess of the algorithm section.

A \emph{kernel} of a set $H$ is a minimal set $K$ generating $H$ via the
relation $\mathord\to = (\to_\mathrm{o} \cup \to_\mathrm{p} \cup
\to_\mathrm{c})$, i.e.,
\begin{align*}
  H = \set{ h' : \exists h \in K.\ h \to^\ast h' }.
\end{align*}
An \emph{abstract data type (ADT)} $A$ is the kernel of the set $H(\mathcal{I})$
of histories of some implementation $\mathcal{I}$, and $H(A)$ denotes the
closure of $A$ under $\to$. While the kernels of arbitrary-width history sets
may not be unique, those with \emph{atomic} kernels, i.e.,~of width-$1$, are
guaranteed to be unique. Section~\ref{sec:nature} demonstrates that the
histories of naturally-occurring implementations have bounded-width and
unique ADTs, which we assume for the remainder of this work.

We define \emph{observational refinement} between implementations and ADTs as
history-set inclusion. Although this refinement is typically defined with
respect to the admissibility of program executions, recent work demonstrates
that these definitions are equivalent~\cite{conf/popl/BouajjaniEEH15}.

@@

  An implementation $\mathcal{I}_1$ \emph{refines} another implementation
  $\mathcal{I}_2$ when $H(\mathcal{I}_1) \subseteq H(\mathcal{I}_2)$. An
  implementation $\mathcal{I}$ refines an ADT $A$ when $\mathcal{I}$ refines
  $H(A)$.

@@

Recent works demonstrate efficient algorithms\footnote{In time polynomial in
the number of operations, per execution.} for checking observational
refinement~\cite{conf/popl/BouajjaniEEH15, conf/pldi/EmmiEH15}, yet rely on
hand-written symbolic ADT representations. In order to frame the problem of
computing these automatically, we fix a language for symbolic representation. A
\emph{history formula} is a first-order logic formula with
\begin{itemize}

@@

  The \emph{symbolic ADT inference problem} is to compute a history formula
  representing the ADT $\ker H(\mathcal{I})$ of a given (reference) implementation
  $\mathcal{I}$.

@@

>>>>

Let $H^*$ denote the \emph{closure} of a history set $H$ under the relation
$\mathord\to = (\to_\mathrm{o} \cup \to_\mathrm{p} \cup \to_\mathrm{c})$, i.e.,
\begin{align*}
  H^* = \set{ h' : \exists h \in H.\ h \to^\ast h' }.
\end{align*}
A set $K$ \emph{generates} $H$ when $K^* = H$, and a \emph{kernel} of a set $H$
is a minimal set $K$ which generates $H$. While the kernels of arbitrary sets of
histories need not be unique, the kernels of sets which have sequential kernels
are unique. Furthermore, Section~\ref{sec:nature} demonstrates that the
histories of naturally-occurring implementations have unique kernels, which we
assume for the remainder of this work. An \emph{abstract data type (ADT)} $A$ is
the kernel of the set $H(\mathcal{I})$  of histories of some implementation
$\mathcal{I}$. We say that \emph{$A$ is the ADT of $\mathcal{I}$} when $A^* =
H(\mathcal{I})$.

ADTs and reference implementations serve as specifications to more efficient
implementations in the sense that they limit the set of histories that efficient
implementation may admit. This notion of refinement ensures that client program
(safety) properties which hold using the ADT or reference implementation also
hold using refined implementations~\cite{conf/popl/BouajjaniEEH15}.

@@

  An implementation $\mathcal{I}_1$ \emph{refines} another implementation
  $\mathcal{I}_2$ when $H(\mathcal{I}_1) \subseteq H(\mathcal{I}_2)$. An
  implementation $\mathcal{I}$ refines an ADT $A$ when $H(\mathcal{I})
  \subseteq A^*$.

@@

Recent works demonstrate efficient\footnote{In time polynomial in the number of
operations, per execution.} refinement-checking
algorithms~\cite{conf/popl/BouajjaniEEH15, conf/pldi/EmmiEH15} yet rely on
hand-written symbolic ADT representations. In order to frame the problem of
computing these automatically, we fix a language for symbolic representation. A
\emph{history formula} is a first-order logic formula with
\begin{itemize}

@@

  The \emph{symbolic ADT inference problem} is to compute a history formula
  representing the ADT of a given implementation.

@@

Computing a history formula representing the ADT of a reference implementation
$\mathcal{I}$ thus enables efficient modular program reasoning without the
burden of writing precise formal specifications for $\mathcal{I}$.
